draft 1

### 2016-06-15 finalwriteup/draft.rmd
reviewed motivation

B2B - Business to Business 
CRM - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.

@@

Statistical research projects on reference forecasting have shown promising results in IT, construction, and film projects. However, persuading industry to adopt these methods is a separate challenge.

** you are here **

#### Industry Uptake of Existing Models and Current Practises

Surveys on current industry practise in cost estimation are scarce; a survey of the construction industry in the UK in 2000 found that a manager's experience and intuition is still the dominant method for construction cost calculation [@Akintoye2000]. A study on the IT industry by @Jorgensen2007 cited that the parametric models were not comprehensive enough to be used in industry and also found that expert judgment was not mobilised enough in the process of developing these tools.

Although reference class models have not gained traction in either the software or construction industry (where most of the research to date has focussed), it is worth outlining their current practises in cost estimation and the reasons behind these.

In the construction industry, infrastructure and building projects follow a traditional and well established process, where contractual norms have developed over many decades [@Badenfelt2011]. The industry is so large, broad and well established that it has inertia in the way that projects are estimated, and contractually bound. The contractor (builder) first reviews the drawings and if the plans are not detailed enough, they offer a temporary budget price. Once they feel that the drawings are sufficiently detailed, they will offer a fixed price [@Badenfelt2011]. However, as the job progresses, if any work is added that was not in the detailed drawings, they maintain the right to charge a variation [@Badenfelt2011]. This method is clear and variations are easily justifiable to the client, however cost, labour and time estimates are still regularly under calculated, even from detailed drawings. And for large infrastructure projects, small percentage miscalculations can be exceedingly expensive.

Contractor's practise of pricing from detailed drawings and charging for variations does not translate to other consultants in the construction industry, such as engineers and architects. Consultants must devlop the detailed plans for the contractors, and must therefore offer fixed price contracts based on drawings with little detail. It is much harder for a consultant to charge a variation for a portion of their time spent on something that was not included in the original simple plans. This is despite the real and common instance of budget blow outs due to overspent time [@Harris1999]. One method to provide opportunity for a variation is stating a minimum percentage fee of the final building cost, which can overrule the original fixed price contract. If a consultant has spent time beyond their fixed price budget,this may account for some or in the best case, all, of their overspent time.

@@

In contrast to construction, IT projects are a relatively new practise where alternative contractural arangements have been trialled. Through experimentation with contracts, the Agile method was developed. In summary, the method treats both cost and time as fixed quantities for a project, and if any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted ad revised based on frequent progress review points that assesses unexpected problems. This agile attitude has become ingrained in many IT practises and this flexible upfront arrangement with clients often forges better long term relationships [Cockburn2001].

It is not completely clear why industry uptake of cost estimation tools using the outside view has been weak, and it may be the case that in modern industries such as in the IT industry, problems associated with cost estimation can be innovatively designed out of the contract. Possible reasons that parametric reference class forecasting models developed by researchers have not been incorporated into industry practise include:

* lack of appropriate data
    * if a project is unique, such as a national infrastructure project, it many not feel appropriate to gather sufficient data points from other countries and time periods.
* lack of understanding from industry decision makers - the prediction process was not explained well in marketing of the tools
* models were not developed in collaboration with industry
* does not feel intuitive price from an 'outside view' of a project which disregards small project details
* insufficient time allocated to pricing that makes change burdensome

In fact, a study by @Moores1992 concluded that a lack of framework to support the outside view model use and a failure in marketing were the two most common explanations for lack of uptake of cost estimating tools which use statistical techniques in predicting project costs. This is unfortunate as almost all research indicates an outside view analysis of a job substantially improves cost estimation accuracy for fixed price projects.

@@

This research will explore the potential for reference class forecasting to improve cost estimation in the context of an engineering consulting company in the construction industry. In this field, labour is the chief cost and the long traditional history of this industry mandates fixed price projects as the norm. In recent years, the size of losses from unprofitable projects in the case study could equal up to 25-30% of the profits from profitable projects. This demonstrates substantial room for improvement and the possiblity of multitude positive flow on effects.

Cost estimation is currently performed by first carefully reviewing preliminary drawings for a project where time costs will be calculated using a manager's personal experience with similar projects. This will often be cross-checked with a value based off a percentage of the estimated final cost of construction (generally the client budget). No formal mathematical method of comparison to similar projects is performed. This is the case for a number of possible reasons:

* lack of time to interrogate the CRM
* clunky data availability offered by the CRM
* lack of awareness of projects performed by other managers that may have been similar.

Reference class forecasting in this study differs from previous research, in that statistical analysis will be performed on internal past company project data that has been progressively recorded in the company's Customer Relationship Management (CRM) software which was purchased. This presents an opportunity to evaluate reference class forecasting in a unique situation, where the 'reference' projects are internal and over two thousand cases are available.

The CRM data currently stands as an untapped source of information within the case study organisation. It stores a rich variety of data including:

@@

A statistical model predicting a measure of project financial success will be developed using the available project variables and trialling numerous machine learning algorithms and statistical methods. The company uses well established CRM software which readily provides simple output statistics such as project time-cost vs invoiced amount summaries. However, the analytical capabilities were limited to:

* simple scatter plots and bar charts of the raw data
* summaries such as overall hours spent vs invoiced amount for each project

The company has a wealth of data but limited means to extract insight. This study will exploit the CRM data by performing more sophisticated statistical analysis with the intention of building a predictive model for cost estimation. The benefits of this study in comparison to previous cases are that thousands of past data points are available, when previous studies used on average 15-20 cases and at most 170. This improves the potential for accurate predictions. If successful, the model also has higher potential for managerial uptake as the algorithm will be directly built and trained on company data. Managers can interrogate 'similar' data points and relate to the actual projects or speak to a colleague who was involved. This has the potential to give the 'outside view' stronger influence on the final decision.

@@

The case study provides a framework to test the value of a consulting company's CRM data in assisting project cost estimates. However, it is important to be aware of limitations when reviewing outcomes and framing the problem. The most obvious limitation is that we are disection a single company in a specific industry, which limits the ability to state conclusions could apply to a broad range of cases. This can be overcome with future studies.

The data set from this case study does not include information that would improve project cost predictions. For example, managers commonly price jobs based on the budget for the entire project, of which the consultant is a minor part. The manager might typically moderate his fee by say 1% of the entire project cost. The value of the entire project has typically not been recorded although there is a space in the project software form for this value. It would be even more beneficial for managers to record the expected total project cost at the beginning of a project versus the conclusion of a completed construction project. Another piece of data that was not recorded was external factors influencing the fixed price quote including competition from similar firms that the client played against the case study company.

In addition to this data that was not typically entered by workers, the dataset also excludes detailed information about the projects which are crucial to the final pricing. This includes details such as preliminary drawings describing the project (square meterage, number of storeys), and project summaries provided by the client to enable a quote. This kind of detailed information will never be recorded in a CRM and in many ways contradicts the purpose of the 'outside view' which intends to present similar projects without being over-influenced by the finer details of a project. Given the lack of detailed information, it is not reasonable for an algorithm to predict the specifc fee of a project. It is much more effective to attempt predict the profitability of a project (ie return per dollar) based on the historical data. This profitability prediction relates to the profitability of the project if the fee was generated using the established method the company used in its historical cases.

Finally, CRM data is entered by employees manually as time passes, jobs are invoiced, and projects are finally completed. Because the data is input by humans, it is susceptible to input error. Many of errors were detectable during data cleaning, however it is possible that undetectable human errors exist in the data.

Although this case study presents an opportunity to test the predictive power held within the organisations' existing CRM data, limitations exist which must be understood and built into the framework for predictions. These include missing data that could have been entered for each project, but historically has not been required, missing data that is very detailed and will never be available, and the potential for a small degree of human error to be embedded within the data.

@@

Fixed price cost estimation for complex projects is a difficult task and when performed incorrectly, can have devastating affects on businesses and governments. Many theories have delved into human nature and our tendency to optimistically assess our capabilities with respect to a given task. In this case, the optimism often results in negative business consequences affecting employee morale and the capacity of a company to flourish and grow. Over the past few decades, significant research has been dedicated to creating predictive models that present the 'outside view' of a project, by statistically comparing a new project to numerous similar projects and their characteristics. It has been shown that these models improve the cost estimation, however industry uptake has not been successful. It has proven to be a challenging task to persuade managers to use this so called reference class models in practise and remove themselves from additions based on intricate details of the completed project. This may be due to a number of reasons including lack of understanding, time, and lack of collaboration with industry in development of these models. This research examines the potential for a company's internally generated CRM data to be used in creating a reference class forecasting model. This has the potential to influence the managers' decisions in the company more however its limitations must be understood in order to create a good model.

@@

**General Aim** Use statistical techniques to model the profitability of projects for consulting businesses using their internal CRM data. Research will focus on a case study Engineering consulting company that offers their expert advice (in the currency of time) to business clients. The project outcomes are intended to assist the business in  predicting project costs before project engagement and other business analyses such as client analysis. Several statistical and machine learning techniques will be tested, compared and refined.

@@

**Hypothesis 2**
The predictive model built from Hypothesis 1 can be proven to have a positive impact on the bottom line of the case study business.

## Thesis Contributions - Gaps in literature

* a gap in the study of predicting internal performance in businesses with business clients. The literature has assessed performance of movies, customer analytics but rarely employee performance or business clients. This is particularly relevant to consulting companies that tackle complicated discrete projects with set timelines. Statistical and machine learning models to predict the correct fee for projects according to project type, client, and other characteristics could reduce risk for a range of consulting businesses in setting initial fees.

* A gap in machine learning predictive models which cater for variables with large amounts of missing data. the nature of consulting jobs is quite complex and historically, not all information about a project is possible to obtain. Variables with missing data are alternatively treated as optional bonus variables in a multi-step model as opposed to missing information which must be imputed or deletion of the entire case that contains missing information.

@@

This chapter provides an overview of the literature available about current business to business (B2B) cost estimation methods, use cases of CRM data, as well as machine learning algorithms and statistical business prediction models. The most prominent statistical and machine-learning models developed for business applications shall be reviewed along with models applied to the most closely related business cases to this project. Gaps in past research will also be highlighted.

@@

This project will tackle the problem of estimating fixed fees for complex consulting projects. It is therefore imperitive to understand what research has been performed to date and how this project contributes to the body of work. The task of estimating fixed fees can be likened to the terms 'cost estimation' or 'effort estimation' in the literature. The bulk of research to date has been performed with project data in either the construction industry or software development. The research into cost estimation in each of these industries is reviewed below, detailing different methods and outcomes. Finally a gap in the literature will be highlighted which this research aims to fill.

@@

Research into cost estimation in the construction industry chiefly focuses on predicting the final building cost of construction. Various studies have surveyed popular methods for cost estimation while others have  investigated new methods such as building information modelling (BIM) analyisis, case based reasoning, and predictive algorithms. The cost estimation methods of detailed analysis (inside view), case based reasoning, and predictive algorithms shall be reviewed.

@@

Detailed analysis refers to the process of an engineer or builder in the construction team carefully reviewing construction drawings to sum the cost of materials, labour, machinery hire, overheads, profit etc. [@Akintoye2000].  @Akintoye2000's study based in the UK found that this is the most popular method for pricing amongst contractors. The second most popular method was to perform an additional manual comparison with similar projects and make judgments based on personal experience. Note that this does not involve a mathematical model.

@Elfaki2014 discusses how much detailed cost estimates for cosntruction projects can vary from estimator (engineer) to estimator, which contributes to the lack of accuracy in predciting final cost of a project. He argues that an engineer's expertise is not documented or measured in any way and therefore their expertise and, in turn, their estimate is prone to subjectivity. @Shane2009 theorises that final cost can be subject to so many different unpredictable parameters, such as weather, unexpected ground conditions, duration or sub-contractor issues that it is almost impossible to achieve an accurate cost prediction manually.

Groundbreaking methods into calculating construction costs by @Ma2014 mobilise the data from building information modelling, although this method is still technically a detailed analysis of the costs. The idea is to quantify the cost of building a structure directly from a three dimensional model created by the design consultants (engineers and architects). @Ma2014's first trial was to automate the cost calculation of a reinforced concrete structure. His chief development was to program his algorithm to intelligently establish construction techniques for each element as this is not provided by architectural or engineering models. From the construction technique and material, a cost can be determined for each element. This was successful and Ma aim's to further his work so that more construction information can be intelligently obtained from BIM models. A system like Ma's would definitely accelerate the cost estimation process and reduce human error, however it is still an 'inside view' and could be prone to innaccurate estimates in a similar way to a detailed analysis by a person. The detailed view does not 'step back' and take into account set backs in construction that may have affected similar projects in the past. There is potential for this to be built into the same system but this is still in early stages of development.

@@

As the problem of cost estimation in construction projects is well recognised, dozens of studies have built prediction models of these projects using machine learning and statistical models. The aim is to test whether a mathematical comparison of many similar projects, using meaningful descriptive variables, can predict the cost of construction. This section will first review the previous studies and most accurate algorithms and finally summarise industry uptake of the methods.

@Elfaki2014 completed a study on cost estimation techniques and research from 2004 to 2014. It was found that artificial neural networks and support vector machines were the most common machine learning techniques used. He highlighted that these two machine learning techniques deal well with uncertainty, however lack technical justification to the decision maker (known as a black box). Neural networks are also time consuming to train, and must be re-trained and re-tested with each additional piece of data [@Kim2004]. Nevertheless, neural networks and SVM's recieved a significant amount of attention in the 1990's for their ability to accurately predict construction costs with limited detailed information [@Shin2015][@Kim2004].

@Shin2015 pioneered the application of boosted trees to cost estimation in construction projects. This is surprising given the rapid uptake of boosted trees over the past decade. In the study, data from 234 school building construction projects in Korea was used. Boosted trees were compared to neural networks, which are one of the most popular algorithms to apply to the problem. Boosted trees were found to predict costs more slightly accurately than neural networks, but not significantly (p value < 0.5). Even though the two algorithms performed similarly, the output from boosted tree models provide insight into the structure of the model such as variable importance and variable partial dependence plots. This additional insight may mean boosted trees are preferable over neural networks because of their potential to engage decision makers.

A notable amount of literature has studied the predictive power of multiple linear regression in the construction cost estimation problem. Often, linear regression is the only model assessed, without comparison to other methods such as neural networks, which first started appearing in literature in the 1990's [@Kim2004]. However, even post-1990, many studies focused on linear regression only. This may be because the technique is straightforward, easy to use and widely available in statistical packages [@Chan2005]. Out of those that have compared regression to neural networks, the studies have shown that neural networks outperform regression, however other studies show they are approximately equal [@Kim2004][@Attalla2003]. @Dissanayaka1999 found that neural networks outperformed multiple linear regression models, however regression models could be performed first to determine variable importance and condense the number of variables. It can be argued that boosted trees could perform as well as neural networks while providing additional insights (even more so than regression models) such as variable importance and partial dependence of variables against cost.

Despite neural networks typically outperforming linear regression models, it is intersting to note some of the outcomes of regression studies in cost estimation. Different studies tend to produce a wide range of variables that contribute to different degrees in the models. This is likey due to the nature of construction projects which are multidisciplinary and involve a range of different parties such as the clients, consultants, contractors and suppliers, each of which can have a range of characteristics and degrees of input [@Chan2005]. @Chan2005's study on 87 building projects in Singapore included special, complex projects. Therefore, variables such as contractor's specialised skills, who the client was (public vs private), the client level of experience, and the contractor's financial management highly influenced final cost of the project. Other studies found variables such as project complexity, duration, team experience, information availablility, site requirements, and labour climate to contribute most [@Akintoye2000][@Trost2003][@Pinto1988]. As these studies obtained information from questionaires, the available variables also largely depended on the questions in the questionaires, and therefore each study produces a range of different important variables. Despite the differences in qualitative outcomes, the studies all generally report their mean absolute error or a similar metric from the tested models. This is encouraging, however there is a gap in the literature in how to translate the improved errors from the model into an argument for industry to uptake the model as an improvement on their bottom line.

>>>>

B2B - Business to Business 
CBR - Case based reasoning. A method of estimating by using the results of similar cases.
CRM - Customer Relationship Management. Refers to a common type of software used by businesses to record client and project details as well as employee timesheet records. Generally accessable to each employee in a company.
IT - Information technology
BIM - Building Information Modelling. A digital representation of physical characteristics of an object or structure.

@@

Statistical research projects on reference forecasting have shown promising results in IT, construction, and film projects. However, persuading industry to adopt these methods into everyday business is a separate challenge.

#### Industry Uptake of Existing Models and Current Practices

A survey of the construction industry in the UK in 2000 found that a manager's experience and intuition is still the dominant method for construction cost calculation [@Akintoye2000]. Surveys on current industry practise in cost estimation are scarce, however another study on the IT industry by @Jorgensen2007 argued that previously developed parametric models were not comprehensive enough to be used in industry and that expert judgment was not mobilised enough in the process of developing these tools.

Although the surveys reveal reference class models have not gained traction in either the software or construction industry, it is worth outlining their current practices in cost estimation, fee structures, and the reasons behind these. In the construction industry, infrastructure and building projects follow a traditional process, where contractual norms have developed over many decades [@Badenfelt2011]. The industry is so large, broad and well established that it has inertia in the way projects are estimated, and contractually bound. The contractor (builder) first reviews the drawings and if the plans are not detailed enough, they offer a temporary budget price. Once they feel that the drawings are sufficiently detailed, they will offer a fixed price [@Badenfelt2011]. As the job progresses however, if additional work on top of the detailed drawings is required, the builder maintains the right to charge a variation [@Badenfelt2011]. A dual source of income therefore exists, the initial proposed fee for construction along with variations due to additional materials (and time required to install the materials) that were not explicitly specified on drawings. Variations fees are generally calculated after completion the work, but the initial construction fee contains risk. Material costs are forseeable but labour and time estimates can still be under calculated, even from detailed drawings. For large infrastructure projects, small percentage miscalculations can be exceedingly expensive.

The builders' procedure of calculating a fixed price from detailed drawings and charging for variations does not translate to other consultants in the construction industry involved with the design of structures, such as engineers and architects. These consultants develop the detailed plans for the builders, and must therefore offer fixed price contracts based on low-detail preliminary drawings that sum up the client's expectations. It is also much harder for a consultant to charge variations for portions of their time spent on items that were not included in the original simple plans. Time is less tangible than physical additional materials required, and often is shrugged off by consultants for the sake of client relationships. This is despite the real and common instance of budget blow outs due to overspent time [@Harris1999]. One method of creating a variation opportunity is by stipulating a percentage fee of the final building cost that can overrule the original fixed price contract. If a consultant has spent time beyond their fixed price budget, this may account for some or in the best case, all, of their overspent time.

@@

In contrast to construction, IT projects are a relatively new practise where alternative contractural arangements have been trialled. Through experimentation with contracts, the Agile movement was developed. In summary, the method treats both cost and time as fixed quantities for a project, and any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted ad revised based on frequent progress review points that assesses unexpected problems. This agile attitude has become ingrained in many IT practices and the flexible upfront arrangement with clients often forges better long term relationships [Cockburn2001].

It is not completely clear why industry uptake of cost estimation tools using the outside view has been weak, and it may be the case that in modern industries such as the IT industry, problems associated with cost estimation can be innovatively designed out of the contract. Possible reasons that parametric reference class forecasting models developed by researchers have not been incorporated into industry practice include:

* lack of appropriate data
    * if a project is unique, such as a national infrastructure project, it many not feel appropriate to gather sufficient data points from other countries and time periods.
* lack of understanding from industry decision makers - the prediction process was not explained well in marketing of the tools
* models were not developed in collaboration with industry
* does not feel intuitive to price from an 'outside view' of a project which disregards small project details
* insufficient time allocated to pricing that makes change burdensome

In fact, a study by @Moores1992 concluded that a lack of framework to support use of the outside view model and a failure in marketing were the two most common explanations for lack of uptake of parametric cost estimating tools. This is unfortunate as almost all research in the area indicates an outside view analysis of a job substantially improves cost estimation accuracy for fixed price projects [@Moores1992].

@@

This research explored the potential for reference class forecasting to improve cost estimation in the context of an engineering consulting company in the construction industry. In this field, labour is the chief cost and the long traditional history of this industry mandates fixed price projects as the norm. In recent years, the size of losses from unprofitable projects in the case study equalled up to 25-30% of the profits from profitable projects. This demonstrates substantial room for improvement and the possiblity of multitude positive flow on effects.

Cost estimation is currently performed by first carefully reviewing preliminary drawings for a project where time costs will be calculated using a manager's personal experience with similar projects. This is often cross-checked with a value based off a percentage of the estimated final cost of construction (generally the client budget). No formal mathematical method of comparison to similar projects is performed. This is the case for a number of possible reasons:

* lack of time to interrogate past project data
* clunky data availability offered by the current database software
* lack of awareness of similar projects performed by other managers

A Customer Relationship Management software package (CRM) was employed by this company to collect and store project data. It is available to all employees over the company intranet, and each employee completes daily timesheets alotting their hours to certain projects. Additionally, information is recorded against each project as well as each client, client contact, and each invoice.

Reference class forecasting in this study differs from previous research in that statistical analysis was performed on internal past company project data that had been progressively recorded in the company's CRM. This presented an opportunity to evaluate reference class forecasting in a unique situation, where the 'reference' projects were all internal and a comparatively large number of cases (over 2,000 projects) were available.

The CRM data stood as an untapped source of information within the case study organisation. It stores a rich variety of data including:

@@

A statistical or machine learning model predicting a measure of project financial success was developed using the available project information and by trialling numerous machine learning algorithms and statistical methods. The company uses well established CRM software which readily provides simple output statistics, however, analytical capabilities were limited to:

* simple scatter plots and bar charts of the raw data
* summaries such as overall hours spent vs. invoiced amount for each project

The company had a wealth of data but limited means to extract insight. This study will exploit the CRM data by performing more sophisticated statistical analysis with the intention of building a predictive model for cost estimation. The benefits of this study in comparison to previous cases are that thousands of past data points are available, when previous studies used on average 15-20 cases and at most 300 [@Finnie1997][@Pai2013][@Shin2015]. This improves the potential for accurate predictions. If successful, the model also has higher potential for managerial uptake as the algorithm will be directly built and trained on company data. Managers can interrogate 'similar' data points and relate to the actual projects or speak to a colleague who was involved. This has the potential to give the 'outside view' stronger influence on the final decision.

@@

The case study provided a framework to test the value of a consulting company's CRM data in assisting project cost estimates. However, it was important to be aware of limitations when reviewing outcomes and framing the problem. The most obvious limitation is that a single company is being disected in a specific industry, which limits the ability to state conclusions applicable to a broad range of cases. This can be overcome with future studies. Other constraining factors include various forms of missing data and human data entry error.

The data set from this case study includes scarce information on the overall cost of entire construction projects. For example, managers commonly price jobs based on the budget for the entire project, of which the consultant is a minor part. The manager might typically moderate his fee by say 1% of the entire project cost. The value of the entire project has typically not been recorded although there is an allotted space in the project database for this. It would be even more beneficial for managers to record the expected total project cost at the beginning of a project versus the concluding completed cost. Another piece of data not recorded was the external factors influencing the fixed price quote including competition from similar firms that clients may have played against the case study company.

In addition to the missing data mentioned above that could be entered by workers in the future, the database also excludes the detailed information about projects that contribute to final pricing. Such details include the preliminary drawings describing the project (square meterage, number of storeys), and project summaries provided by the client to enable a quote. This level of information will never be recorded in a CRM and in many ways contradicts the purpose of the 'outside view', the intention of which is to present similar projects without being over-influenced by the finer details. Given this lack of detailed information, it was not reasonable for an algorithm to predict the numeric fee for a project. It appeared much more effective to predict the profitability of a project (ie return per dollar) based on available data. This profitability is a ratio between amount invoiced and business costs. The algorithm therefore would predict the ratio of these two values based on fees and costs generated using the established method of fee proposals and time expenditure in historical cases.

Finally, CRM data is entered by employees manually as time passes, jobs are invoiced, and projects are finally completed. Because the data is input by people, it is susceptible to input error. Many errors were detectable during data cleaning, however it is possible that a low number of undetectable human errors exist in the data and marginally influence predictions.

Although this case study presented an opportunity to test the predictive power held within the organisations' existing CRM data, limitations exist which must be understood and built into the framework of the problem. These include missing data that could have been entered for each project, but historically has not been required, missing data that is very detailed and will never be available, and the potential for a small degree of human error to be embedded within the data.

@@

Fixed price cost estimation for complex projects is a difficult task and when performed incorrectly, can have devastating affects on businesses and governments. Many theories have delved into human nature and our tendency to optimistically assess our capabilities with respect to a given task. In this case, the optimism often results in negative business consequences affecting employee morale and the capacity of a company to flourish and grow. Over the past few decades, significant research has been dedicated to creating predictive models that present the 'outside view' of a project, by statistically comparing a new project to a collection of similar projects and their characteristics. It has been shown that these models improve the cost estimation, however industry uptake has not been successful. It has proven to be a challenging task persuading managers to use reference class forecasting models in practise and remove themselves from summaries based on intricate details of the completed project. This may be due to a number of reasons including lack of understanding, time, and lack of collaboration with industry in development of these models. This research examines the potential for a company's internally generated CRM data to be used in creating a reference class forecasting model. Familiarity with the data has the potential to influence managers' decisions in the company more, however its limitations must be understood in order to create a valid model.

@@

**General Aim** Use statistical techniques to model the profitability of projects for consulting businesses using their internal CRM data. Research will focus on a case study Engineering consulting company that offers their expert advice (in the currency of time) to business clients.

The project outcomes are intended to assist the business in  predicting project costs before project engagement and other business analyses such as client analysis. Several statistical and machine learning techniques will be tested, compared and refined.

@@

**Hypothesis 2**
The predictive model built from Hypothesis 1 can be shown to have a positive impact on the bottom line of the case study business.

## Thesis Contributions

This work contributes to the body of research in numerous ways. There is a gap in the literature of applying statistical and machine learning methods to internally generated business data to facilitate internal business decisions. Studies have modelled the commercial performance of movies, customer churn, the share market, and bankruptcies but few have modelled more day-to-day data that a business generates which could actually assist day-to-day decisions. Improving business decisions would reduce risk and improve profits in scenarios that could be applicable in thousands of businesses around the world.

In the case of this study, the problem of cost estimation is tackled. This issue has been worked on extensively in past research, however the problem has rarely been addressed using a single company's internal data. In the literature, dozens of projects from other companys and even countries are collected and analysed. This thesis uses the internal CRM data from a company which contains over 2,000 past job cases - far more than the usual cost estimation study. Furthermore, the problem of cost estimation is applied to a consulting company in the construction industry that estimates and must stand by their fixed fees before a project is won. This differs from other studies estimating actual construction costs of projects and is likened more to software development effort estimation although the nature of the work differs considerably.

In addition to addressing cost estimation with a different kind of dataset and different industry sector, ensemble tree methods will be applied which have been minimally tested on this problem. To date, linear regression, neural networks, case based reasoning, and SVM's have been used to predict cost/effort even though ensemble tree methods can perform as well as neural networks in some cases and generally outperform linear regression. Another benefit of ensemble trees is the output types available, such as partial dependency plots and variable importance plots provide insight into the model's calculations. This is in contrast to neural networks and support vector machines (SVM's) which are 'black-box' predictors. Providing insight to a user is a major benefit in this case as past research has demonstrated issues with user trust and uptake of models, even if the models were successful. It is anticipated that using internal data only also improves user trust and creates tangible meaning for users. Blending multiple machine learning and statistical models was trialled which has not yet been applied to the cost estimation problem.

Finally, this study takes the model one step further than previous studies and analyses concrete profit improvements for the company if the model output were to be used in certain decision-making scenarios. A few possibilities are presented as ways to take model results into account with business decisions and where possible, final profit calcuations are presented. It is anticipated that this kind of thought experiment would further push industry to seriously consider the worth in adopting statistical cost estimation models.

## Thesis Structure

The remainder of this thesis has the following structure: the next chapter is a thorough review of the literature relating to this thesis to date. This covers work solving the cost-estimation problem, studies on statistical and machine-learning models in business applications, and a review of effective statistical and machine learning methods.

The following chapter, Chapter 4, describes the method followed throughout the research process. It includes first how the data was obtained, the lengthy cleaning process, followed by variable importance analysis, variable selection, trials of selected predictive algorithms and problem constructs. The cost estimation problem was attempted first as a regression problem predicting return per dollar, followed by a simpler binary classification problem predicitng profit or loss. Once the best methods were selected, they were blended in numerous ways, using both simple averaging techniques and sophisticated machine-learning algorithms. These were compared against individual models and the best constructs were selected. Finally the impact of the algorithm on the overall profits of the case study company was analysed via decision making scenarios and further applications were considered.

Chapter 5 presents the results of the series of analyses described above and directly relates them to the overall aim and hypotheses of the project. This is in the format of tables and plots where results are briefly discussed. The examination of results is taken deeper as well as in a broader context in Chapter 6: Discussion. Here the nature of the results are debated in terms of how surprising they were, their impact on the research hypothesis and limitations as well as suggested future work. The final Chapter 7: Conclusion summarises the  findings of the project, how the research answered the hypotheses and how the work could be viably applied in industry.

@@

This chapter provides an overview of the literature available in current cost estimation methods, use cases of CRM data, as well as machine learning algorithms and statistical business prediction models. The most prominent statistical and machine-learning models developed for business applications shall be reviewed along with models applied to business cases that can be related to this project. Gaps in past research will also be highlighted.

@@

The application for this project is the task of estimating fixed fees for complex consulting projects. It is therefore prudent to understand what research has been performed to date and how this project contributes to the body of work. The task of estimating fixed fees can be likened to the terms 'cost estimation' or 'effort estimation' in the literature. The bulk of research to date has been performed with project data in either the construction industry or software development. Research into cost estimation in each of these industries is reviewed below, detailing different methods and outcomes. Finally a gap in the literature will be highlighted which this research aims to fill.

@@

Research on cost estimation in the construction industry has primarily focused on predicting the final building cost of construction. Various studies have surveyed methods for cost estimation while others have investigated new methods such as building information modelling (BIM) analyisis, case based reasoning, and predictive algorithms. The cost estimation methods of detailed analysis (inside view), case based reasoning, and predictive algorithms are reviewed in this section.

@@

Detailed analysis refers to the process of an engineer or builder in the construction team carefully reviewing construction drawings to sum the cost of materials, labour, machinery hire, overheads, and profit etc. [@Akintoye2000].  @Akintoye2000's study based in the UK found that this is the most popular method for pricing amongst contractors. The second most popular method was to carry out an additional comparison with similar projects and make judgments based on personal experience. Note that this does not involve a mathematical model.

@Elfaki2014 discusses how much detailed cost estimates for construction projects can vary from estimator (engineer) to estimator, which contributes to the lack of accuracy in predciting final cost of a project. He argues that an engineer's expertise is not documented or measured in any way and therefore their expertise and, in turn, their estimate is prone to subjectivity. @Shane2009 theorises that final cost can be subject to so many different unpredictable parameters, such as weather, unexpected ground conditions, duration or sub-contractor issues, that it is almost impossible to achieve an accurate cost prediction manually.

Groundbreaking methods into calculating construction costs by @Ma2014 mobilise the data from BIM models, although this method is still technically a detailed analysis of the costs. The idea is to quantify the cost of building a structure directly from a three dimensional model created by the design consultants (engineers and architects). @Ma2014's first trial was to automate the cost calculation of a reinforced concrete structure. His chief contribution was to program his algorithm to intelligently establish construction techniques for each element as this is not provided by architectural or engineering models. From the construction technique and material, a cost can be derived for each element. This was successful and @Ma2014 aims to further his work so that more construction information can be intelligently obtained from BIM models. A system like @Ma2014's would definitely accelerate the cost estimation process and reduce human error, however it is still an 'inside view' and could be prone to innaccurate estimates similarly to detailed analysis by a person. The detailed view does not 'step back' and take into account set backs in construction that may have affected similar projects in the past. There is potential for this innovation to be combined with an outside-view system but @Ma2014's is still in early stages of development.

@@

As the problem of cost estimation in construction projects is well recognised, dozens of studies have built predictive models of these projects using machine learning and statistical models. The aim is to test whether a mathematical comparison of many similar projects, using meaningful descriptive variables, can predict the cost of construction. This section will first review the previous studies and most accurate algorithms and finally summarise industry uptake of the methods.

@Elfaki2014 completed a study on cost estimation techniques and research from 2004 to 2014. They found that artificial neural networks and SVM's were the most common machine learning techniques used. He highlighted that these two machine learning techniques deal well with uncertainty, however lack technical justification to the decision maker (known as a black box). Neural networks are also time consuming to train, and must be re-trained and re-tested with each additional piece of data [@Kim2004]. Nevertheless, neural networks and SVM's recieved a significant amount of attention in the 1990's for their ability to accurately predict construction costs with limited detailed information [@Shin2015][@Kim2004].

@Shin2015 pioneered the application of boosted trees to cost estimation in construction projects. This is surprisingly late given the rapid uptake of boosted trees over the past decade. In the study, data from 234 school building construction projects in Korea were used. Boosted trees were compared to neural networks, which are one of the most popular algorithms to apply to this problem. Boosted trees were found to predict costs slightly more accurately than neural networks, but not statstically significantly (p value < 0.5). Even though the two algorithms performed similarly, the output from boosted tree models provide insight into the structure of the model such as variable importance and variable partial dependence plots. This additional insight may mean boosted trees are preferable over neural networks because of their potential to engage decision makers.

A notable amount of literature has studied the predictive power of multiple linear regression in the construction cost estimation problem. Often, linear regression is the only model assessed, without comparison to other methods such as neural networks, which first started appearing in literature in the 1990's [@Kim2004]. However, even post-1990, many studies focused on linear regression only. This may be because the technique is straightforward, easy to use and widely available in statistical packages [@Chan2005]. Out of those that have compared regression to neural networks, the studies have shown that neural networks outperform regression, however other studies established they are approximately equal [@Kim2004][@Attalla2003]. @Dissanayaka1999 found that neural networks outperformed multiple linear regression models, however regression models could be performed first to determine variable importance and condense the number of variables. It can be argued that boosted trees could perform as well as neural networks while providing additional insights (even more so than regression models) such as variable importance and partial dependence of variables against the independent variable, cost.

Despite neural networks typically outperforming linear regression models, it is interesting to note some of the outcomes of regression studies in cost estimation. Different studies tended to produce a wide range of variables that contribute at fluctuating levels in the models. This is likely because construction projects are multidisciplinary and involve a range of different parties such as the clients, consultants, contractors and suppliers, each of which contribute varying levels of input [@Chan2005]. @Chan2005's study on 87 building projects in Singapore included special, complex projects. Therefore, variables such as contractor's specialised skills, who the client was (public vs. private), the client level of experience, and the contractor's financial management highly influenced final cost of the project. Other studies found variables such as project complexity, duration, team experience, information availablility, site requirements, and labour climate to contribute most [@Akintoye2000][@Trost2003][@Pinto1988]. As these studies obtained information from questionaires, the available variables also largely depended on the questions in the questionaires, and therefore each study produced a range of different important variables. Despite the differences in qualitative outcomes, the studies all generally reported their mean absolute error or a similar metric from the tested models. Similar result structures are useful, however there is a gap in the literature in how to translate the improved errors from the model into a well justified argument for industry to reduce losses by integrating a parametric model into their business.

Surveys into construction industry uptake of these algorithmic models are sparse, however a survey by @Akintoye2000 showed that parametric estimating methods have not been adopted by contractors, despite the research into this area over the 20 prior years. The majority of contractors use experience based methods. @Akintoye2000 hypothesises that this could be due to a lack of familiarity and knowledge about the techniques, doubts whether these techniques are applicable to the construction industry, and the availability of sound data to ensure confidence. This presents a clear gap in the literature on how to assess the models in terms of the end user's goals and limitations.
